---
title: Reading the eDNA dataset
output: md_document
editor_options: 
  chunk_output_type: console
---

# Reading the eDNA dataset

## Reading occurrences

Read the individual Occurrence and DNADerivedData files, and join them:

```{r read, message=FALSE, warning=FALSE}
library(dplyr)
library(stringr)
library(purrr)

dna_files <- list.files("../output", "*DNADerivedData*", full.names = TRUE)
occurrence_files <- list.files("../output", "*Occurrence*", full.names = TRUE)

dna <- map(dna_files, read.table, sep = "\t", quote = "", header = TRUE) %>%
  bind_rows() %>%
  mutate_if(is.character, na_if, "")

occurrence <- map(occurrence_files, read.table, sep = "\t", quote = "", header = TRUE) %>%
  bind_rows() %>%
  mutate_if(is.character, na_if, "") %>%
  mutate(
    species = ifelse(taxonRank == "species", scientificName, NA),
    aphiaid = as.numeric(str_replace(scientificNameID, "urn:lsid:marinespecies.org:taxname:", ""))
  ) %>%
  left_join(dna, by = "occurrenceID")
```

## Resolving to accepted species

The PacMAN pipeline aligns taxa with WoRMS, but this may included unaccepted taxa such as synonyms. Use the procedure below to resolve all taxa to their accepted names. This can take a few minutes.

```{r resolve, message=FALSE, warning=FALSE}
library(worrms)
library(furrr)

resolve_to_accepted <- function(occurrence) {
  aphiaids <- unique(occurrence$aphiaid)
  aphiaid_batches <- split(aphiaids, as.integer((seq_along(aphiaids) - 1) / 50))
  plan(multisession, workers = 4)
  aphiaid_mapping <- future_map(aphiaid_batches, wm_record) %>%
    bind_rows() %>%
    select(aphiaid = AphiaID, valid_aphiaid = valid_AphiaID) %>%
    distinct() %>%
    filter(aphiaid != valid_aphiaid)
  
  valid_aphiaids <- unique(aphiaid_mapping$valid_aphiaid)
  valid_aphiaid_batches <- split(valid_aphiaids, as.integer((seq_along(valid_aphiaids) - 1) / 50))
  valid_taxa <- map(valid_aphiaid_batches, wm_record) %>%
    bind_rows() %>%
    select(valid_aphiaid = AphiaID, scientificName = scientificname, scientificNameID = lsid, taxonRank = rank, kingdom, phylum, class, order, family, genus) %>%
    mutate(taxonRank = tolower(taxonRank))
occurrence %>%
  mutate(verbatimScientificName = scientificName) %>%
  left_join(aphiaid_mapping, by = "aphiaid") %>%
  rows_update(valid_taxa, by = "valid_aphiaid") %>%
  mutate(
    species = ifelse(taxonRank == "species", scientificName, NA),
    aphiaid = as.numeric(str_replace(scientificNameID, "urn:lsid:marinespecies.org:taxname:", ""))
  ) %>%
  select(-valid_aphiaid)
}

occurrence <- resolve_to_accepted(occurrence)
```

Inspect updated names:

```{r updatednames, message=FALSE, warning=FALSE}
occurrence %>%
  filter(scientificName != verbatimScientificName) %>%
  group_by(scientificName, verbatimScientificName) %>%
  summarize(n = n()) %>%
  arrange(desc(n)) %>%
  head(10) %>%
  knitr::kable()
```

## Data exploration

```{r exploration, message=FALSE, warning=FALSE, fig.width=15, fig.height=12, fig.dpi=300}
library(ggplot2)

stats <- occurrence %>%
  filter(!is.na(phylum) & !is.na(higherGeography)) %>%
  group_by(higherGeography, phylum) %>%
  summarize(n = n_distinct(species))

ggplot() +
  geom_bar(data = stats, aes(y = higherGeography, x = n, fill = phylum), stat = "identity") +
  viridis::scale_fill_viridis(discrete = TRUE) +
  theme_minimal() +
  theme(legend.position = "bottom", axis.title = element_blank())
```
